### 2025-07-23 paper/sections/other/constraints.qmd
formatting improved and toc added (to this end, turned appendix into standalone document).

In @sec-constraints we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature. We argue in @sec-constraints that this approach induces models to be less sensitive to immutable features and demonstrate this empirically in @sec-experiments. Below we derive the analytical results in Prp.~\ref{prp-mtblty}.

@@

Let $\mathbf{x}^\prime$ denote some randomly chosen individual from class $y^-$ and let $y^+ \sim p(y)$ denote the randomly chosen target class. Then the partial derivative of the contrastive divergence penalty [@eq-div] with respect to coefficient $\theta_{y^+,d}$ is equal to

>>>>

In the main paper, we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature. We argue that this approach induces models to be relatively less sensitive to immutable features, propose a theoretical result supporting this and provide empirical evidence that strengthens our argument (both in the main paper and additional findings in this appendix). Below we derive the analytical results in Prp.~\ref{prp-mtblty}.

@@

Let $\mathbf{x}^\prime$ denote some randomly chosen individual from class $y^-$ and let $y^+ \sim p(y)$ denote the randomly chosen target class. Then the partial derivative of the contrastive divergence penalty with respect to coefficient $\theta_{y^+,d}$ is equal to

### 2025-07-23 paper/sections/other/counterfactuals.qmd
trying to address comments on figure 2
formatting improved and toc added (to this end, turned appendix into standalone document).

For all generators used in this work we use standard logit crossentropy loss for $\text{yloss}(\cdot)$. All generators also penalize the distance ($\ell_1$-norm) of counterfactuals from their original factual state. For *Generic* and *ECCo*, we have $\mathcal{Z}:=\mathcal{X}$ and $g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}$, that is counterfactual are searched directly in the feature space. Conversely, *REVISE* traverses the latent space of a variational autoencoder (VAE) fitted to the training data, where $g(\cdot)$ corresponds to the decoder [@joshi2019realistic]. In addition to the distance penalty, *ECCo* uses an additional penalty component that regularizes the energy associated with the counterfactual, $\mathbf{x}^\prime$ [@altmeyer2024faithful].

>>>>

For all generators used in this work we use standard logit crossentropy loss for $\text{yloss}(\cdot)$. All generators also penalize the distance ($\ell_1$-norm) of counterfactuals from their original factual state. For *Generic* and *ECCCo*, we have $\mathcal{Z}:=\mathcal{X}$ and $g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}$, that is counterfactual are searched directly in the feature space. Conversely, *REVISE* traverses the latent space of a variational autoencoder (VAE) fitted to the training data, where $g(\cdot)$ corresponds to the decoder [@joshi2019realistic]. In addition to the distance penalty, *ECCCo* uses an additional penalty component that regularizes the energy associated with the counterfactual, $\mathbf{x}^\prime$ [@altmeyer2024faithful].

<!-- TODO: Add note on why we omit the conformal prediction component. -->

### 2025-07-23 paper/sections/other/grids.qmd
formatting improved and toc added (to this end, turned appendix into standalone document).

1. **Generator Parameters** (@sec-app-grid-gen): Investigates the effect of changing hyperparameters that affect the counterfactual outcomes during the training phase.
2. **Penalty Strengths** (@sec-app-grid-pen): Investigates the effect of changing the penalty strengths in out proposed objective (@eq-obj).
3. **Other Parameters** (@sec-app-grid-train): Investigates the effect of changing other training parameters, including the total number of generated counterfactuals in each epoch.

@@

To measure predictive performance, we compute the accuracy and F1-score for all models on test data (@tbl-acc-gen, @tbl-acc-pen, @tbl-acc-train). With respect to explanatory performance, we report here our findings for the (im)plausibility and cost of counterfactuals at test time. Since the computation of our proposed divergence metric (@eq-impl-div) is memory-intensive, we rely on the distance-based metric for the grid searches. For the counterfactual evaluation, we draw factual samples from the training data for the grid searches to avoid data leakage with respect to our final results reported in the body of the paper. Specifically, we want to avoid choosing our default hyperparameters based on results on the test data. Since we are optimizing for explainability, not predictive performance, we still present test accuracy and F1-scores.

@@

We do observe strong sensitivity to certain hyperparameters, with clear an manageable patterns. Concerning generator parameters, we firstly find that using *REVISE* to generate counterfactuals during training typically yields the worst outcomes out of all generators, often leading to a substantial decrease in plausibility. This finding can be attributed to the fact that *REVISE* effectively assigns the task of learning plausible explanations from the model itself to a surrogate VAE. In other words, counterfactuals generated by *REVISE* are less faithful to the model that *ECCo* and *Generic*, and hence we would expect them to be a less effective and, in fact, potentially detrimental role in our training regime. Secondly, we observe that allowing for a higher number of maximum steps $T$ for the counterfactual search generally yields better outcomes. This is intuitive, because it allows more counterfactuals to reach maturity in any given iteration. Looking in particular at the results for *Linearly Separable*, it seems that higher values for $T$ in combination with higher decision thresholds ($\tau$) yields the best results when using *ECCo*. But depending on the degree of class separability of the underlying data, a high decision-threshold can also affect results adversely, as evident from the results for the *Overlapping* data (@fig-grid-gen_params-plaus-over): here we find that CT generally fails to achieve its objective because only a tiny proportion of counterfactuals ever reaches maturity.

@@

imgfname = "plausibility_distance_from_target.png"
fig_caption = "Average outcomes for the plausibility measure across hyperparameters. This shows the % change from the baseline model for the distance-based implausibility metric (@eq-impl-dist). Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCo*)."
full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption, width=img_width) 
_str = join(include_img_commands, "\n\n")
println(_str)
```

@@

imgfname = "distance.png"
fig_caption = "Average outcomes for the cost measure across hyperparameters. This shows the % change from the baseline model for the distance-based cost metric [@wachter2017counterfactual]. Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCo*)."
full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption, width=img_width)
_str = join(include_img_commands, "\n\n")
println(_str)
```

>>>>

1. **Generator Parameters** (@sec-app-grid-gen): Investigates the effect of changing hyperparameters that affect the counterfactual outcomes during the training phase.
2. **Penalty Strengths** (@sec-app-grid-pen): Investigates the effect of changing the penalty strengths in our proposed training objective.
3. **Other Parameters** (@sec-app-grid-train): Investigates the effect of changing other training parameters, including the total number of generated counterfactuals in each epoch.

@@

To measure predictive performance, we compute the accuracy and F1-score for all models on test data (@tbl-acc-gen, @tbl-acc-pen, @tbl-acc-train). With respect to explanatory performance, we report here our findings for the (im)plausibility and cost of counterfactuals at test time. Since the computation of our proposed divergence-based adaption ($\text{IP}^*$) is memory-intensive, we rely on the distance-based metric for the grid searches. For the counterfactual evaluation, we draw factual samples from the training data for the grid searches to avoid data leakage with respect to our final results reported in the body of the paper. Specifically, we want to avoid choosing our default hyperparameters based on results on the test data. Since we are optimizing for explainability, not predictive performance, we still present test accuracy and F1-scores.

@@

We do observe strong sensitivity to certain hyperparameters, with clear an manageable patterns. Concerning generator parameters, we firstly find that using *REVISE* to generate counterfactuals during training typically yields the worst outcomes out of all generators, often leading to a substantial decrease in plausibility. This finding can be attributed to the fact that *REVISE* effectively assigns the task of learning plausible explanations from the model itself to a surrogate VAE. In other words, counterfactuals generated by *REVISE* are less faithful to the model that *ECCCo* and *Generic*, and hence we would expect them to be a less effective and, in fact, potentially detrimental role in our training regime. Secondly, we observe that allowing for a higher number of maximum steps $T$ for the counterfactual search generally yields better outcomes. This is intuitive, because it allows more counterfactuals to reach maturity in any given iteration. Looking in particular at the results for *Linearly Separable*, it seems that higher values for $T$ in combination with higher decision thresholds ($\tau$) yields the best results when using *ECCCo*. But depending on the degree of class separability of the underlying data, a high decision-threshold can also affect results adversely, as evident from the results for the *Overlapping* data (@fig-grid-gen_params-plaus-over): here we find that CT generally fails to achieve its objective because only a tiny proportion of counterfactuals ever reaches maturity.

@@

imgfname = "plausibility_distance_from_target.png"
fig_caption = "Average outcomes for the plausibility measure across hyperparameters. This shows the % change from the baseline model for the distance-based implausibility metric (\$\text{IP}\$). Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCCo*)."
full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption, width=img_width) 
_str = join(include_img_commands, "\n\n")
println(_str)
```

@@

imgfname = "distance.png"
fig_caption = "Average outcomes for the cost measure across hyperparameters. This shows the % change from the baseline model for the distance-based cost metric [@wachter2017counterfactual]. Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCCo*)."
full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption, width=img_width)
_str = join(include_img_commands, "\n\n")
println(_str)
```

### 2025-07-23 paper/sections/other/initial_grids.qmd
formatting improved and toc added (to this end, turned appendix into standalone document).

- **Energy Penalty** (@tbl-lin_sep-lambda_energy_exper): *ECCo* generally does yield better results than *Vanilla* for higher choices of the energy penalty (10,15) during training. *Generic* performs poorly accross the board. *Omni* seems to have an anchoring effect, in that it never performs terribly but also never as good as the best *ECCo* results. *REVISE* performs poorly across the board.
- **Cost** (@tbl-lin_sep-lambda_cost_exper): Results for all generators (except *Omni*) are quite bad, which can likely be attributed to extremely bad results for some choices of the **Energy Penalty** (results here are averaged). For *ECCo* and *Generic*, higher cost values generally lead to worse results.
- **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
- **Validity**: *ECCo* almost always valid except for very low values during training and high values at evaluation time. *Generic* often has poor validity.
- **Accuracy**: Seems largely unaffected.

@@

- **Energy Penalty** (@tbl-moons-lambda_energy_exper): *ECCo* consistently yields better results than *Vanilla*, except for very low choices of the energy penalty during training for which it performs abismal. *Generic* performs quite badly across the board for high enough choices of the energy penalty at evaluation time. *Omni* has small positive effect. *REVISE* performs poorly across the board.
- **Cost (distance penalty)**: *Generic* generally does better for higher values, while *ECCo* does better for lower values.
- **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
- **Validity**: *ECCo* generally achieves full validity except for very low choices the energy penalty during training and high choices at evaluation time. *Generic* performs poorly for high choices of the energy penalty during evaluation.
- **Accuracy**: Largely unaffected although *ECCo* suffers a bit for very low choices the energy penalty during training. *REVISE* suffers a lot in general (around 10 percentage points).

@@

- **Energy Penalty** (@tbl-circles-lambda_energy_exper): *ECCo* consistently yields better results than *Vanilla*, though primarily for low to medium choices of the energy penalty (<=5) during training. The same goes for *Generic*, which sometimes outperforms *ECCo* (for small energy penalty at evaluation time). *Omni* does alright for lower energy penalty at evaluation time, but loses out for higher choices. *REVISE* performs poorly across the board (except very low choices at evaluation time).
- **Cost (distance penalty)**: *ECCo* and *Generic* generally achieve the best results when no cost penalty is used during training. Both *Omni* and *REVISE* are largely unaffected.
- **Maximum Iterations**: *ECCo* consistently yields better results for higher numbers of iterations. *Generic* generally does best for a medium number (50). *Omni* is sometimes invalid (**???**).
- **Validity**: *ECCo* tends to outperform its *Vanilla* counterpart, though primarily for low to medium choices of the energy penalty (<=5) during training and evaluation. *Vanilla* typically worse across the board.
- **Accuracy**: Mostly unaffected, but *REVISE* again consistently some deterioration and *ECCo* deteriorates for high choices of energy penalty during training, reflecting other outcomes above.

>>>>

- **Energy Penalty** (@tbl-lin_sep-lambda_energy_exper): *ECCCo* generally does yield better results than *Vanilla* for higher choices of the energy penalty (10,15) during training. *Generic* performs poorly accross the board. *Omni* seems to have an anchoring effect, in that it never performs terribly but also never as good as the best *ECCCo* results. *REVISE* performs poorly across the board.
- **Cost** (@tbl-lin_sep-lambda_cost_exper): Results for all generators (except *Omni*) are quite bad, which can likely be attributed to extremely bad results for some choices of the **Energy Penalty** (results here are averaged). For *ECCCo* and *Generic*, higher cost values generally lead to worse results.
- **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
- **Validity**: *ECCCo* almost always valid except for very low values during training and high values at evaluation time. *Generic* often has poor validity.
- **Accuracy**: Seems largely unaffected.

@@

- **Energy Penalty** (@tbl-moons-lambda_energy_exper): *ECCCo* consistently yields better results than *Vanilla*, except for very low choices of the energy penalty during training for which it performs abismal. *Generic* performs quite badly across the board for high enough choices of the energy penalty at evaluation time. *Omni* has small positive effect. *REVISE* performs poorly across the board.
- **Cost (distance penalty)**: *Generic* generally does better for higher values, while *ECCCo* does better for lower values.
- **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
- **Validity**: *ECCCo* generally achieves full validity except for very low choices the energy penalty during training and high choices at evaluation time. *Generic* performs poorly for high choices of the energy penalty during evaluation.
- **Accuracy**: Largely unaffected although *ECCCo* suffers a bit for very low choices the energy penalty during training. *REVISE* suffers a lot in general (around 10 percentage points).

@@

- **Energy Penalty** (@tbl-circles-lambda_energy_exper): *ECCCo* consistently yields better results than *Vanilla*, though primarily for low to medium choices of the energy penalty (<=5) during training. The same goes for *Generic*, which sometimes outperforms *ECCCo* (for small energy penalty at evaluation time). *Omni* does alright for lower energy penalty at evaluation time, but loses out for higher choices. *REVISE* performs poorly across the board (except very low choices at evaluation time).
- **Cost (distance penalty)**: *ECCCo* and *Generic* generally achieve the best results when no cost penalty is used during training. Both *Omni* and *REVISE* are largely unaffected.
- **Maximum Iterations**: *ECCCo* consistently yields better results for higher numbers of iterations. *Generic* generally does best for a medium number (50). *Omni* is sometimes invalid (**???**).
- **Validity**: *ECCCo* tends to outperform its *Vanilla* counterpart, though primarily for low to medium choices of the energy penalty (<=5) during training and evaluation. *Vanilla* typically worse across the board.
- **Accuracy**: Mostly unaffected, but *REVISE* again consistently some deterioration and *ECCCo* deteriorates for high choices of energy penalty during training, reflecting other outcomes above.

### 2025-07-23 paper/sections/other/main_results.qmd
formatting improved and toc added (to this end, turned appendix into standalone document).

As discussed @sec-experiments, CT is sensitive to certain hyperparameter choices. We study the effect of many hyperparameters extensively in @sec-app-grid. For the main results, we tune a small set of key hyperparameters (@sec-app-tune). The final choices for the main results are presented for each data set in @tbl-final-params along with training, test and batch sizes.

@@

Final hyperparameters used for the main results presented in @sec-experiments. Any hyperparameter not shown here is set to its default value (@nte-train-default).

@@

Mean outcomes for **CT** and **BL** along with bootstrapped confidence intervals (99%) for difference in mean outcomes grouped by dataset and evaluation metric. Column **LB** and **UB** show the lower and upper bound of the intervals, respectively, and computed using the percentile method. The underlying counterfactual evaluations are the same as the ones used to produce @tbl-main.

@@

::: {layout="[10,-2,10]" layout-valign="top"}
![Counterfactual images for *MLP* with counterfactual training. Factual images are shown on the diagonal, with the corresponding counterfactual for each target class (columns) in that same row. The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp.png){#fig-mnist}

@@

@@

### Costs

@@

### Validity

@@

>>>>

As discussed the main paper, CT is sensitive to certain hyperparameter choices. We study the effect of many hyperparameters extensively in @sec-app-grid. For the main results, we tune a small set of key hyperparameters (@sec-app-tune). The final choices for the main results are presented for each data set in @tbl-final-params along with training, test and batch sizes.

@@

Final hyperparameters used for the main results presented in the main paper. Any hyperparameter not shown here is set to its default value (@nte-train-default).

@@

Mean outcomes for **CT** and **BL** along with bootstrapped confidence intervals (99%) for difference in mean outcomes grouped by dataset and evaluation metric. Column **LB** and **UB** show the lower and upper bound of the intervals, respectively, and computed using the percentile method. The underlying counterfactual evaluations are the same as the ones used to produce the main table in the paper.

@@

::: {layout="[10,-2,10]" layout-valign="top"}
![Counterfactual images for *MLP* with counterfactual training. Factual images are shown on the diagonal, with the corresponding counterfactual for each target class (columns) in that same row. The underlying generator, *ECCCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp.png){#fig-mnist}

@@

::: {layout-ncol=2}

@@

![Class-conditional integrated gradients.](/paper/figures/mnist_ig.png){#fig-mnist-ig}

### Costs and Validity

::: {#tbl-panel layout-ncol=3}

@@

@@

Costs and validity.

### 2025-07-23 paper/sections/other/tuning.qmd
formatting improved and toc added (to this end, turned appendix into standalone document).

::: {.callout-warning}

@@

@@

imgfname = "plausibility_distance_from_target.png"
fig_caption = "Average outcomes for the plausibility measure across key hyperparameters. This shows the % change from the baseline model for the distance-based implausibility metric (@eq-impl-dist). Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCo*)."
full_paths = joinpath.(data_dirs, joinpath(ce_suffix,imgfname))
include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
_str = join(include_img_commands, "\n\n")
println(_str)
```

>>>>

<!-- ::: {.callout-warning}

@@

::: -->

@@

imgfname = "plausibility_distance_from_target.png"
fig_caption = "Average outcomes for the plausibility measure across key hyperparameters. This shows the % change from the baseline model for the distance-based implausibility metric (\$\text{IP}\$). Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCCo*)."
full_paths = joinpath.(data_dirs, joinpath(ce_suffix,imgfname))
include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
_str = join(include_img_commands, "\n\n")
println(_str)
```
